Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-03-07 numerics_paper/version2/numerical_chj.tex
Work the remarks more naturally into the text.

\section{Numerical methods for the shallow water equations} \label{sec2}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + \nabla \cdot f(q) & = 0,
\end{align}
where $q=[h, hu, hv]^T$ and the flux functions $f$ is defined in accordance
with \eqref{eq:sw}.

@@

\subsection{The entropy residual}
Let $\entvar(q)$ and $\entflux(q)$ denote the entropy variable and entropy flux.
Recall that $q=[h,hu,hv]^T$. We use the total energy as entropy; that is, 
\begin{align*}
  \eta(q)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(q)=
    \frac{\eta(q)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(q) \cdot \left[ \frac{\partial q}{\partial t} + \nabla\cdot \bff(q)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(q)}{\partial t} + \entvar(q) \cdot \nabla\cdot \bff(q)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(q)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(q) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(q) \cdot \nabla\cdot \bff(q) - \nabla\cdot\entflux(q) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $q$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\entvar(\bfq_i)\cdot \int_{S_i}\nabla\cdot \bff(q)d\bfx
  -\int_{S_i}\nabla\cdot\entflux(q) d\bfx\right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(q)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\entflux(q)d\bfx\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

Let $\entvar_i=\entvar(\bfq_i)$ and $\entflux_i=\entflux(\bfq_i)$ denote the entropy variable and the
(one-dimensional) entropy flux at cell $i$, respectively.
Also let $\efp=\entvar(\bfq_i)\cdot \bff(\bfq_i)-\entflux(\bfq_i)$ be the entropy potential at cell $i$.
From \cite[\S 4]{tadmor1987numerical}, if the numerical fluxes satisfy
\begin{align}\label{es_cond}
(\entvar_{i}-\entvar_{i-1})\cdot \bfF_{i-1/2}\leq \efp_{i}-\efp_{i-1},
  \qquad
  (\entvar_{i+1}-\entvar_i)\cdot \bfF_{i+1/2}\leq \efp_{i+1}-\efp_i,
\end{align}
then the following entropy inequality holds
\begin{align}\label{ent_ineq}
  \frac{d\eta(\bfq_i)}{dt}+\frac{1}{\Delta x}\left[G_{i+1/2}-G_{i-1/2}\right]\leq 0,
\end{align}
where
\begin{align*}
    G_{i+1/2} &= \frac{1}{2}\left(\entvar_i+\entvar_{i+1}\right)\bfF_{i+1/2}-\frac{1}{2}(\efp_{i}+\efp_{i+1}),
\end{align*}
is a discretization of the entropy flux.

After plugging \eqref{blended-flux-1} into the condition \eqref{es_cond}, we arrive at the value
\begin{align}\label{lambda_min}
  \lambda_{i+1/2}^{\min} 
  = \max\left\{0,\frac{\frac{1}{2}(\entvar_{i+1}-\entvar_{i})\cdot
    \left[\bff(\bfq_{i+1})+\bff(\bfq_{i})-\sum_p\lambda_{i+1/2}^{p,\EV}\W_{i+1/2}^{p,\Roe}\right]
    -(\efp_{i+1}-\efp_i)}
  {\frac{1}{2}(\entvar_{i+1}-\entvar_{i})\cdot\sum_p\W_{i+1/2}^{p,\Roe}}\right\},
\end{align}
where 
\begin{align*}
  \lambda_{i+1/2}^{p,\EV}=\theta_{i+1/2}\lambda_{i+1/2}^{\max}+(1-\theta_{i+1/2})|\hat{\lambda}_{i+1/2}^p|.
\end{align*}
With this choice, \eqref{blended-flux-1} guarantees \eqref{es_cond}, and therefore \eqref{ent_ineq}.
Note that here we have used the identities
\begin{align*}
|\hat{A}_\iph|(Q_{i+1}-Q_i) = \sum_p |\hat{\lambda}^p_\iph| \W^{p,\Roe}_\iph, \qquad
Q_{i+1}-Q_i & = \sum_p \W^{p,\Roe}_\iph.
\end{align*}

@@

\begin{remark}
In terms of fluctuations, the entropy stability condition \eqref{es_cond} can be written as
\begin{align*}
  (\entvar_{i}-\entvar_{i-1})\cdot
  \Big(
  \underbrace{\bff(\bfq_i)-\A^+\Delta Q_{i-1/2}}_{=\bfF_{i-1/2}}\Big)\leq \efp_{i}-\efp_{i-1},
  \quad
  (\entvar_{i+1}-\entvar_i)\cdot
  \Big(\underbrace{\bff(\bfq_i)+\A^-\Delta Q_{i+1/2}}_{=\bfF_{i+1/2}}\Big)\leq \efp_{i+1}-\efp_i.
\end{align*}
\end{remark}

\begin{remark}[About the effect of $\lambda_{i\pm 1/2}^{\min}$ for Roe's solver.]
  If we set $\theta_{i\pm 1/2}=0$ and $\lambda^{\min}_\iph=0, ~\forall i$ in \eqref{lambda_p}, then the Riemann solver becomes Roe's solver.
  In this case, it is well known that transonic rarefactions develop non-physical shocks,
  a phenomenon known as `entropy glitch'. To fix this problem, many entropy fixes to Roe's
  solver have been developed. Enhancing the dissipation via \eqref{lambda_min}
  also serves as a transonic entropy fix for Roe's solver, which we demonstrate
  via a numerical experiment in \S \ref{sec:rp_dry_bed}.
\end{remark}

\begin{remark*}[About the design philosophy.]
  The work by Tadmor \cite{tadmor2003entropy} is based on imposing condition \eqref{ent_ineq} as equality. 
  By doing this, the method is said to be entropy conservative. To improve the nonlinear 
  stability properties of the numerical scheme, a consistent (and potentially high-order) 
  artifical entropy dissipative operator is included. 
  In this work, we add dissipation on the conserved variables based on 
  a measurement of the entropy residual 
  (as in \cite{guermond2011entropy,guermond2018second,guermond2018well}).
  Doing this, however, does not guarantee \eqref{ent_ineq}. 
  Indeed, some linear stabilization techniques that add artificial dissipation on the conserved 
  variables are known to produce entropy; see for example \cite{ern2013weighting,kuzmin2020entropy}. 
  To guarantee \eqref{ent_ineq}, we add extra dissipation on the conserved variables via 
  \eqref{lambda_min}. 
  Doing this counteracts entropy production created by any component of the Riemann solver;
  see for example \cite{kuzmin2020entropy} (in the context of $C^0$ finite elements). 
\end{remark*}

>>>>

\section{Numerical methods for the shallow water equations} \label{sec2}
We consider the shallow water model in two horizontal dimensions:
\begin{subequations} \label{eq:sw}
\begin{align}
    h_t + (hu)_x + (hv)_y & = 0 \\
    (hu)_t + \left(hu^2 + \frac{1}{2}gh^2\right)_x + (huv)_y & = 0 \\
    (hv)_t + (huv)_x + \left(hv^2 + \frac{1}{2}gh^2\right)_y & = 0.
\end{align}
\end{subequations}
Here $h, u,$ and $v$ are respectively the depth and the $x$- and $y$-components of
velocity, which are functions of the two spatial coordinates $(x,y)$ as well as time $t$.
The gravitational force is proportional to $g$.
The system \eqref{eq:sw} can be written in vector form as
\begin{align}
    q_t + \nabla \cdot f(q) & = 0,
\end{align}
where $q=[h, hu, hv]^T$ and the flux function $f$ is defined in accordance
with \eqref{eq:sw}.

@@

\subsection{The entropy residual}
Let $\entvar(q)$ and $\entflux(q)$ denote the entropy variable and entropy flux.
Recall that $q=[h,hu,hv]^T$. We use the total energy as entropy; that is, 
\begin{align*}
  \eta(q)=\frac{1}{2} gh^2 + \frac{(hu)^2+(hv)^2}{2h}, \qquad
  \entflux(q)=
    \frac{\eta(q)}{h}
    \begin{bmatrix} hu \\ hv \end{bmatrix}.
\end{align*}
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \entvar(q) \cdot \left[ \frac{\partial q}{\partial t} + \nabla\cdot \bff(q)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(q)}{\partial t} + \entvar(q) \cdot \nabla\cdot \bff(q)\right] d\bfx
\end{align}
as a measurement of the entropy production in cell $i$ (here $S_i=[x_\imh,x_\iph]$).
To avoid the need to compute the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(q)}{\partial t}d\bfx=-\int \nabla\cdot\entflux(q) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\entvar(q) \cdot \nabla\cdot \bff(q) - \nabla\cdot\entflux(q) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound on $R_i$ so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $q$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \Delta x \left|\entvar(Q_i)\cdot \nabla\cdot f(Q_i)
  -\nabla\cdot\entflux(Q_i) \right|,
  \quad
  %D_i = ||\entvar(\bfq_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfq)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfq)d\bfx\right|.
  D_i = \Delta x \sum_{k=1}^{d+1}|\entvar_k(\bfq_i)|\left|\left(\nabla\cdot\bff(Q_i)\right)_k\right|
  +\left|\nabla\cdot\entflux(Q_i)\right|,
\end{align*}
where $(z)_k$ denotes the $k$-th component of $z\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions.  In \eqref{blended-flux-1} we need values of $\theta$ at the cell interfaces, for
which we use
$$
    \theta_\iph = \max(\theta_i, \theta_{i+1}).
$$
We expect that $\theta_\iph\approx 0 $ in smooth regions,
while $\theta_\iph\approx 1$ near shocks.
The value of $\theta_\iph$ controls whether the first-order flux \eqref{blended-flux-1}
behaves more like that of Rusanov or Roe.  Specifically, if
$\lambda_\iph^{\min}=0$ and $\theta_\iph=1$, the flux is equivalent to that of
Rusanov, while if $\lambda_\iph^{\min}=0$ and $\theta_\iph=0$, it is equivalent
to that of Roe.

@@

Let $\entvar_i=\entvar(\bfq_i)$ and $\entflux_i=\entflux(\bfq_i)$ denote the entropy variable and the
(one-dimensional) entropy flux at cell $i$, respectively.
Also let $\efp=\entvar(\bfq_i)\cdot \bff(\bfq_i)-\entflux(\bfq_i)$ be the entropy potential at cell $i$.
From \cite[\S 4]{tadmor1987numerical}, if the numerical fluxes satisfy
\begin{align}\label{es_cond}
(\entvar_{i}-\entvar_{i-1})\cdot \bfF_{i-1/2}\leq \efp_{i}-\efp_{i-1},
  \qquad
  (\entvar_{i+1}-\entvar_i)\cdot \bfF_{i+1/2}\leq \efp_{i+1}-\efp_i,
\end{align}
then we have the entropy inequality
\begin{align}\label{ent_ineq}
  \frac{d\eta(\bfq_i)}{dt}+\frac{1}{\Delta x}\left[G_{i+1/2}-G_{i-1/2}\right]\leq 0,
\end{align}
where
\begin{align*}
    G_{i+1/2} &= \frac{1}{2}\left(\entvar_i+\entvar_{i+1}\right)\bfF_{i+1/2}-\frac{1}{2}(\efp_{i}+\efp_{i+1}),
\end{align*}
is a discretization of the entropy flux.
If \eqref{ent_ineq} holds with equality, the scheme is entropy-conservative
\cite{tadmor2003entropy}.

The added dissipation described in the previous sections does not guarantee \eqref{ent_ineq}. 
Indeed, some linear stabilization techniques that add artificial dissipation of the conserved 
variables are known to produce entropy; see for example \cite{ern2013weighting,kuzmin2020entropy}. 
To guarantee \eqref{ent_ineq}, we add extra dissipation of the conserved variables via 
\eqref{lambda_min}. 
Doing this counteracts entropy production created by any component of the Riemann solver;
see for example \cite{kuzmin2020entropy} (in the context of $C^0$ finite elements).

Plugging \eqref{blended-flux-1} into the condition \eqref{es_cond} yields the required value
\begin{align}\label{lambda_min}
  \lambda_{i+1/2}^{\min} 
  = \max\left\{0,\frac{\frac{1}{2}(\entvar_{i+1}-\entvar_{i})\cdot
    \left[\bff(\bfq_{i+1})+\bff(\bfq_{i})-\sum_p\lambda_{i+1/2}^{p,\EV}\W_{i+1/2}^{p,\Roe}\right]
    -(\efp_{i+1}-\efp_i)}
  {\frac{1}{2}(\entvar_{i+1}-\entvar_{i})\cdot\sum_p\W_{i+1/2}^{p,\Roe}}\right\},
\end{align}
where 
\begin{align*}
  \lambda_{i+1/2}^{p,\EV}=\theta_{i+1/2}\lambda_{i+1/2}^{\max}+(1-\theta_{i+1/2})|\hat{\lambda}_{i+1/2}^p|.
\end{align*}
With this choice, \eqref{blended-flux-1} guarantees \eqref{es_cond}, and therefore \eqref{ent_ineq}.
Note that here we have used the identities
\begin{align*}
|\hat{A}_\iph|(Q_{i+1}-Q_i) = \sum_p |\hat{\lambda}^p_\iph| \W^{p,\Roe}_\iph, \qquad
Q_{i+1}-Q_i & = \sum_p \W^{p,\Roe}_\iph.
\end{align*}

@@

We close this section with two remarks.  First, note that 
the entropy stability condition \eqref{es_cond} can be written equivalently in terms of fluctuations:
\begin{align*}
  (\entvar_{i}-\entvar_{i-1})\cdot
  \Big(
  \underbrace{\bff(\bfq_i)-\A^+\Delta Q_{i-1/2}}_{=\bfF_{i-1/2}}\Big)\leq \efp_{i}-\efp_{i-1},
  \quad
  (\entvar_{i+1}-\entvar_i)\cdot
  \Big(\underbrace{\bff(\bfq_i)+\A^-\Delta Q_{i+1/2}}_{=\bfF_{i+1/2}}\Big)\leq \efp_{i+1}-\efp_i.
\end{align*}

Second, the additional dissipation introduced by $\lambda_{i\pm 1/2}^{\min}$ can also
serve independently as an entropy fix for Roe's solver, as we demonstrate
via a numerical experiment in \S \ref{sec:rp_dry_bed}.
